---
title             : "Examining side effect variability of antipsychotic \
                    treatment in schizophrenia spectrum disorders: A \
                    meta-analysis of variance"
shorttitle        : "Variability of side effects"

author: 
  - name          : "Maria S. Neumeier"
    affiliation   : "1" 
  - name          : "Stephanie Homan, Ph.D."
    affiliation   : "1"
  - name          : "Stefan Vetter, M.D."
    affiliation   : "1" 
  - name          : "Erich Seifritz, M.D."
    affiliation   : "1" 
  - name          : "John M. Kane, M.D."
    affiliation   : "2,3,4" 
  - name          : "Maximilian Huhn, M.D."
    affiliation   : "5"
  - name          : "Stefan Leucht, M.D."
    affiliation   : "5"
  - name          : "Philipp Homan, M.D., Ph.D."
    affiliation   : "1,2,3,4*"
    corresponding : yes    # Define only one corresponding author
    address       : "University Hospital of Psychiatry Zurich, \
                      Lenggstrasse 31, \
                      8032 Zurich, Switzerland" 
    email         : "philipp.homan@bli.uzh.ch"

affiliation:
  - id            : "1"
    institution   : "University Hospital of Psychiatry Zurich, \
                     Zurich, Switzerland."
  - id            : "2"
    institution   : "Center for Psychiatric Neuroscience, \
                     Feinstein Institute for Medical Research, \
                     Manhasset, NY, USA."                    
  - id            : "3" 
    institution   : "Division of Psychiatry Research, \
                     Zucker Hillside Hospital, Northwell Health, \
                     New York, NY, USA."
  - id            : "4"
    institution   : "Department of Psychiatry, \
                     Zucker School of Medicine at Northwell/Hofstra, \
                     Hempstead, NY, USA."
  - id            : "5"
    institution   : "Department of Psychiatry and Psychotherapy, \
                     Technical University of Munich, \
	                   School of Medicine, \
                     Munich, Germany."


abstract: |
  Side effects of `r source("sideeffects_do.R")` antipsychotic 
  drugs play a key role in non-adherence of treatment in schizophrenia 
  spectrum disorders (SSD). While clinical observations suggest 
  that side effect variability between patients may be considerable, 
  statistical evidence is 
  required to confirm this. Here, we hypothesized to find larger side effect 
  variability under treatment compared with control. We included double-blind, 
  placebo-controlled, randomized controlled trials (RCTs) of adults with 
  a diagnosis of SSD treated with 1 out of 14 antipsychotics. 
  Standard deviations of the pre-post treatment differences of weight 
  gain, prolactin levels, and corrected QT (QTc) times were extracted. 
  The outcome measure was the variability ratio of treatment to control for 
  individual antipsychotic drugs and the overall variability ratio of 
  treatment to control across RCTs. Individual variability ratios were 
  weighted by the inverse-variance method and entered into a 
  random-effects model. We included N = `r ntot[3]` patients for weight 
  gain, N = `r ntot[1]` patients for prolactin levels, and N = `r ntot[2]` 
  patients for QTc time. Variability ratios (VR) were significantly
  increased for weight gain (`r parse_vrstat(vrstat[[3]])`) and
  prolactin levels (`r parse_vrstat(vrstat[[1]])`) but did not
  reach significance for QTc time (`r parse_vrstat(vrstat[[2]])`).
  We found marked differences between individual antipsychotics and 
  increased variability in side effects in patients under treatment 
  with antipsychotics suggesting that subgroups of patients or 
  individual patients may benefit from treatment allocation through 
  stratified or personalized medicine.


bibliography  : ["sideeffects.bib", "references.bib"]
csl           : nature.csl

leftheader    : ""
fontsize      : 9pt
draft         : no
figsintext    : yes
fig_caption	  : true
figurelist    : no
tablelist     : no
footnotelist  : no
lineno        : no
mask          : no
class         : "jou"            # for pdf change to jou/doc
output        : papaja::apa6_pdf # for pdf change to apa6_pdf/docx
header-includes:
 - \usepackage{libertine}
 - \captionsetup[figure]{font=footnotesize,labelfont=bf}
#- \usepackage[printwatermark]{xwatermark}
#- \usepackage{xcolor}
# - \newwatermark[allpages,color=gray!30,angle=45,scale=3.5,xpos=0,ypos=0]{CONFIDENTIAL}
 - \newcommand{\beginsupplement}{\setcounter{table}{0}
 - \renewcommand{\thetable}{S\arabic{table}}
 - \setcounter{figure}{0}
 - \renewcommand{\thefigure}{S\arabic{figure}}}
#output:
#  papaja::apa6_pdf:
#      draftall
#     citation_package: natbib
# header-includes:
#   - \usepackage[numbered]{natbib}
# bibliostyle   : nature
---
	
```{r global_options, include=FALSE}
# set global options for output images
# tidy.opts = margin cutoff point and wraps text to the next line
knitr::opts_chunk$set(fig.path = "../output/figures/", 
                      echo = FALSE, warning = FALSE, message = FALSE,
                      tidy.opts = list(width.cutoff = 72))

# set global options for knitr
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)

options(tinytex.verbose = TRUE)

```

```{r load_packages, include=FALSE, warning=FALSE, message=FALSE}

fig1 <- "../output/figures/weightsd_fig1.pdf"
fig2 <- "../output/figures/weightsd_fig2.pdf"
fig3 <- "../output/figures/prolactinsd_fig1.pdf"
fig4 <- "../output/figures/prolactinsd_fig2.pdf"
fig5 <- "../output/figures/qtcsd_fig1.pdf"
fig6 <- "../output/figures/qtcsd_fig2.pdf"

# for knitting latex to pdf in windows
Sys.setenv(JAVA_HOME = "C:\\Programme\\Java\\jre1.8.0_171")

```

# Introduction
Antipsychotics are a fundamental component in the treatment of
schizophrenia spectrum disorders (SSD). Yet, a major problem are side
effects which play a key role in non-adherence and
discontinuation.[@Lambert2004; @Kane2013; @Sendt2015; @Wade2017; 
@Homan2019k] A common hypothesis among researchers and clinicians alike 
is that although side effects are pervasive, not all patients are equally
susceptible, even when they are treated with the same drug.[@Huhn2019]
However, empirical support for this hypothesis is lacking, as randomized 
controlled trials (RCTs) or conventional meta-analyses by design cannot 
answer whether such side effect variability does exist.[@Winkelbeiner2019;
@Munkholm2020]

While it is well-established that antipsychotics are associated 
with sides effects for the average patient, the approach we are taking 
with this study moves beyond comparing group averages but instead 
compares group variances. By comparing variances our study can 
for the first time test the hypothesis that there is indeed reason 
to believe that subgroups or even individual patients differ in 
their susceptibility to side effects – something that an analysis 
focused on group averages cannot do.

To date, studies have established the efficacy, safety, and side effect
profiles of antipsychotic medications by averaging these indices across
groups of patients. Such studies can provide us with average side effects 
for specific drugs, but they cannot tell us anything about individual 
patients or subgroups.[@Senn2016; @Senn2018] Nevertheless, before searching 
for potential biomarkers that might predict individual susceptibility, we 
should first quantify the extent to which such predictors are truly needed. 

An approach to answering this question is to shift the focus from the
means to the variances of side effects.[@Cortes2019] By comparing the
variances between treatment and control groups of RCTs,[@Nakagawa2015]
greater variability in treatment would mean that some patients are more
susceptible to side effects than others.[@Cortes2019] Note that this
method[@Mills2020] has recently been applied for 
antipsychotics,[@Winkelbeiner2019] antidepressants,[@Munkholm2020;
@Ploderl2019; @Volkmann2020] and brain stimulation,[@Homan2020a] but in 
the context of treatment effect variability. It is worth noting that these 
studies found little evidence for treatment effect 
variability.[@Winkelbeiner2019; @Munkholm2020; @Ploderl2019; @Volkmann2020] 
Importantly, in the case of pre-post differences used as input for a 
meta-analysis of variance it is crucial to think carefully about the way 
the variability ratio is expressed,[@Nakagawa2015; @Volkmann2020; 
@Volkmann2020a] as the use of the coefficient of variation ratio (CVR) 
that has been proposed as an alternative of the variability ratio 
(VR)[@Nakagawa2015] may lead to unreliable results.[@Mills2020; 
@Volkmann2020a]

A recently published study investigated the individual treatment
response in antipsychotics and brought surprising
results.[@Winkelbeiner2019; @Winkelbeiner2019b] By comparing the
variability between treatment and control groups, no evidence was found
for an increase in variability in the treatment group. What might sound
counter-intuitive at first raises the question of how big the need for
precision medicine really is. However, that study evaluated the evidence
for treatment effect variability. It is possible that although such
variability in treatment effects is not as high as sometimes assumed,
[@Homan2019a] it does exist in the susceptibility for side effects. In
other words, even if there is little variability in response to treatment
between patients, there may still be enough variability in side effects to
justify a need for precision medicine. If true, then this would support
optimization of treatment allocation with respect to side effect
profiles.[@Pillinger2020]

Side effects that are particularly relevant to antipsychotic treatment
include weight gain,[@Homan2019k] hyperprolactinemia, and QTc
prolongation.[@Pillinger2020] Weight gain is a frequently observed side
effect that can negatively impact one's physical health and thus may
also influence treatment adherence. Every additional kilogram of weight
gain can contribute to an increased risk of heart
failure,[@Kenchaiah2002] cardiovascular disease,[@Willett1995] and
diabetes.[@Cooper2016] In addition, treatment discontinuation is often
seen in patients with increase of weight under treatment.[@Mustafa2018]
High prolactin levels can lead to symptoms like decreased bone mass,
gallactorhea, and fertility problems in men and women. Further possible
symptoms include menstrual disturbances in female patients and decreased
libido and erectile dysfunction in male patients.[@Thapa2020] These
symptoms are frequent, but often underreported by patients and unnoticed
as well as untreated by clinicians.[@Montejo2010; @Serretti2011] They
furthermore might lead to loss in quality of life and might be a reason
for treatment discontinuation[@Heald2010; @Lambert2004] and subsequent
illness relapse, which together with persistent positive symptoms
[@Homan2012a; @Cavelti2018; @Cavelti2018a; @Winkelbeiner2018a] may severely
impact recovery and therapeutic alliance.[@Cavelti2016] Prolongation of
QTc was observed in 7 of 14 antipsychotics compared by placebo in the
inter-group comparison by Huhn and colleagues.[@Huhn2019] Importantly,
*torsade de pointes* tachycardia and sudden cardiac death are possible
severe consequences of QTc prolongation.[@Funk2020] Such cardiac events
are one of the factors that lead to the loss of life expectancy observed
in patients with SSD.[@Glassman2005; @Vandael2017; @Koponen2008]

In summary, antipsychotic side effects are highly relevant for
long-term outcome and adherence in treatment of positive symptoms in
SSD.  The question remains whether variability in side effects is high
enough to warrant efforts of treatment stratification or
personalisation. If there is little or no evidence for variability in
side effects there might not be a need for stratification or personalization
and the already widely available data provided by
intergroup comparisons might offer reasonable estimates for the
individual patient. Thus, we compared the variances of side effects
including weight gain, prolactin level and QTc-time between treatment
and control groups to address this question and to evaluate the
evidence for the presence of side effect variability. Based on the
clinical impression that patients seem to vary in their susceptibility
to side effects, we hypothesized that the variability in side effects
would be higher in the treatment compared with the control groups
across all published trials of antipsychotics in SSD.[@Huhn2019]

# Methods
## Search strategy and selection criteria
We used the data from the recent meta-analysis by Huhn and
colleagues.[@Huhn2019] That study included placebo-controlled published
and unpublished trials investigating orally administered atypical
(second generation) antipsychotics and typical (first generation) 
antipsychotics in adults with schizophrenia spectrum disorders; and
excluded patients with first episode psychosis, treatment resistance,
mainly negative symptoms, comorbidity with other mental or physical
illnesses and relapse-prevention studies. Long- and short-acting
intramuscular injections were also excluded (as they are often used in
relapse prevention or emergency treatment) and studies from mainland
China were excluded because of data quality concerns.[@Tong2018]
Data sources were MEDLINE, Cochrane Central Register of Controlled Trials
(CENTRAL), Embase, Biosis, PsycINFO, PubMed, ClinicalTrials.gov, WHO
International Clinical Trials Registry Platform and the US Food and Drug
Administration until January 8 2019. Data quality and validity were
ensured by following the PRISMA guidelines.[@Hutton2015] For missing
data, we also contacted study authors.

We decided to investigate weight gain, prolactin elevation and 
QTc prolongation because these side effects are particularly relevant, 
[@Pillinger2020; @Homan2019k] and quantifiable metric data were 
available in most studies. Other side effects such as 
extrapyramidal motor symptoms (EPS), sedation and diabetes can be 
equally burdensome and certainly pose a health risk to patients. 
However, those have often not been quantified in studies but rather 
assessed qualitatively (e.g. through categorical variables such as 
EPS: yes/no; sedation: yes/no; diabetes: treated/untreated). 
These types of variables did not allow us to estimate variability 
and so we had to restrain from including them in the analysis.

For the analysis, we used the standard deviations of pre-post
differences in side effects. The primary outcome was the 
overall variability ratio of side effects in treatment versus control groups.
Standard deviations (SD) and number of patients (N) were
extracted for weight gain, prolactin level and QTc time. The units used
were kg for weight gain, ng/mL for prolactin levels, and ms for QTc
time. Some studies provided data for all of the three side effects,
whereas the majority of the studies contained less data (see Results).

## Statistical analysis
If patients or subgroups differ in their susceptibility to side effects,
we would expect to observe increased variances in the treatment
compared with the control group. To test this, we computed the log
variability ratio (log VR) by comparing the relative variability of
side effects under treatment versus control:

$$\textrm{log VR}=\textrm{log}\left(\frac{\textrm{SD}_{\textrm{Tx}}}{\textrm{SD}_{\textrm{Ct}}} \right) +
\frac{1}{2(n_{\textrm{Tx}}-1)} - \frac{1}{2(n_{\textrm{Ct}}-1)} $$,

where $\textrm{SD}_{\textrm{Tx}}$ was the reported sample SD for side 
effects under treatment, $\textrm{SD}_{\textrm{Ct}}$ was the reported 
sample SD for side effects under control, $n_{\textrm{Tx}}$ was the 
treatment sample size, and $n_{\textrm{Ct}}$ the control sample size. The 
corresponding sampling variance ($s_{\textrm{log VR}}^2$) for each 
comparison can be expressed as follows:

$$ s_{\textrm{log VR}}^2 = \frac{1}{2(n_{\textrm{Tx}}-1)} + \frac{1}{2(n_{\textrm{Ct}}-1)}$$.

The individual variability ratios were weighted with the inverse of
this sampling variance[@Viechtbauer2010] and entered into a 
random-effects model to quantify the overall variability ratio of side 
effects. For better interpretability, results were back-transformed from 
the logarithmic scale. Here, a variability ratio greater than one would 
indicate a higher side effect variability in treatment compared with 
control, whereas a variability ratio smaller than one would indicate less 
side effect variability under treatment compared with control.

## Data and code availability
The analysis was performed from September 2019 to May 2020, using the
R package metafor[@Viechtbauer2010](version `r packageVersion("metafor")`). 
The manuscript was produced with the R packages rmarkdown 
(version `r packageVersion("rmarkdown")`); represearch 
(version `r packageVersion("represearch")`; 
https://github.com/phoman/represearch/);
knitr (version `r packageVersion("knitr")`); and 
papaja (version `r packageVersion("papaja")`). 
All data and code are freely available online to ensure reproducibility at
https://github.com/homanlab/sideeffects/.

# Results
## Overall reporting details
Together, we screened N = `r norigtrials` studies with 14 different 
antipsychotics from the original meta-analysis by Huhn and 
colleagues[@Huhn2019] as these studies reported data on at least
one of the three side effects that we were interested in. Of these studies,
N = `r sum(!is.na(unique(wgexcl$Study_No)))` 
(`r round(sum(!is.na(unique(wgexcl$Study_No)))/norigtrials, 2) * 100`%) 
had missing variance measures despite reported means for at least 
one of the three side effects. We thus included the N = `r ntrialsincluded` 
(`r round(ntrialsincluded/norigtrials, 2) * 100`%) studies that did 
report variance measures for at least one of the side effects of interest.

## Weight gain 
For weight gain, we included `r nrct[3]` RCTs, yielding `r ncomp[3]` 
comparisons of antipsychotic drugs with placebo to investigate the individual
occurrence of weight gain in patients. All together we included 
N = `r ntot[3]` patients diagnosed with schizophrenia or schizoaffective 
disorder. There were `r ntottx[3]` (`r round(ntottx[3]/ntot[3], 2) * 100`%) 
patients randomly allocated to the treatment group, and 
`r ntotct[3]` (`r round(ntotct[3]/ntot[3], 2) * 100`%) to the placebo
group. Patients in the treatment group received 1 out of 14 
investigated antipsychotic drugs. Individual comparisons between 
drugs across studies indicated marked differences between individual 
antipsychotics. The VR for chlorpromazine, cariprazine, and sertindole was 
smaller than 1. The VR for ziprasidone, iloperidone, and brexpiprazole 
was 1. The VR for asenapine, risperidone, olanzapine, lurasidone, 
aripiprazol, paliperidone, haloperidol, and quetiapine was greater than 1 
(`r parse_vrstat(vrstat[[3]])`; Figure \@ref(fig:fig2)). Overall, the 
variability for weight gain was higher under treatment than
under control (`r parse_vrstat(vrstat[[3]])`; Figure \@ref(fig:fig1)).

```{r fig2capt}
fig2capt <- "Variability ratio for weight gain for individual
antipsychotics. The forest plot shows the VR together with its 95% 
confidence interval (CI) for treatment versus placebo. All included studies 
[@Litman2016;@Garcia2009;@Clark1972;@Kahn2007;
@Ogasa2013;@Davidson2007;@Durgam2016;@Coppola2011;@Landbloom2017;
@Kinon2011;@VanKammen1996;@Arvanitis1997;@Ishigooka2018;
@Correll2015;@Zborowski1995;@Potkin2008;@Nakamura2009;
@Potkin2007;@Keck1998;@Cutler2010;@Clark1970;@Borison1996;@Ahmed2007;
@Durgam2015;@Meltzer2007;@Meltzer2004;@Kane2010;@Kane2002;@Hirayasu2010;
@Daniel1999;@Cooper2000;@Casey2008;@Bugarski2014;@Patil2007;
@Lieberman2015;@Potkin2003;@Durgam2014;@Nasrallah2013;@Marder2007;
@Borison1992;@Shen2014;@Kinoshita2016;@McEvoy2007;@Kane2007;@Harvey2013;
@Meltzer2011;@Canuso2010b;@Potkin2015;@Beasley1996a;@Canuso2010a;
@Beasley1996b;@Loebel2016;@Litman2014;@Kane2016;@Lindenmayer2008]
are also listed in Table S1."

```

```{r fig2, include = TRUE, fig.cap = fig2capt}
knitr::include_graphics(fig2)
```

## Hyperprolactinemia
For hyperprolactinemia, we included `r nrct[1]` RCTs, with `r ncomp[1]`
comparisons of antipsychotic drugs with placebo. All together we
included N = `r ntot[1]` patients diagnosed with schizophrenia or
schizoaffective disorder. There were `r ntottx[1]` 
(`r round(ntottx[1]/ntot[1], 2) * 100`%) patients randomly allocated to 
the treatment group, and `r ntotct[1]` 
(`r round(ntotct[1]/ntot[1], 2) * 100`%) to the placebo group. Patients 
in the treatment group received 1 out of 13 investigated antipsychotic 
drugs. Individual comparisons between drugs across studies indicated 
marked differences between individual antipsychotics. The VR for 
iloperidone and brexpiprazole was smaller than 1. The VR for aripiprazole, 
asenapine, risperidone, cariprazine, sertindole, quetiapine, ziprasidone, 
lurasidone, haloperidol, olanzapine, and paliperidone was greater than 1 
(`r parse_vrstat(vrstat[[1]])`; Figure \@ref(fig:fig4)).
Overall, the variability for hyperprolactinemia was higher under treatment 
than under control (`r parse_vrstat(vrstat[[1]])`; Figure \@ref(fig:fig3)). 

```{r fig4capt}
fig4capt <- "Variability ratio for hyperprolactinemia for individual
antipsychotics. The forest plot shows the VR together with its 95% 
confidence interval (CI) for treatment versus placebo. All included studies
[@Kane2016;@Patil2007;@Kane2015;@Litman2016;
@Loebel2016;@Potkin2008;@Kahn2007;@Correll2015;@Kinoshita2016;
@VanKammen1996;@Small1997;@Kinon2011;@Potkin2003;@Arvanitis1997;
@Durgam2014;@Borison1996;@Schmidt2014;@Meltzer2007;@McEvoy2007;
@Hirayasu2010;@Casey2008;@Beasley1996a;@Marder2007;@Kane2007;
@Kane2010;@Durgam2016;@Lindenmayer2008;@Kane2002;@Ishigooka2018;
@Meltzer2011;@Durgam2015;@Cutler2010;@Harvey2013;@Nasrallah2013;
@Garcia2009;@Canuso2010a;@Nakamura2009;@Geffen2012;@Canuso2010b;
@Zimbroff1997;@Zborowski1995;@Beasley1996b;@Potkin2007;
@Potkin2015;@Davidson2007;@Ogasa2013;@Coppola2011;@Lieberman2015]
are also listed in Table S1." 
```

```{r fig4, include = TRUE, fig.cap = fig4capt}
knitr::include_graphics(fig4)
```


## QTc prolongation
For QTc prolongation, we included `r nrct[2]` RCTs, with `r ncomp[2]`
comparisons of antipsychotic drugs with placebo. All together we
included N = `r ntot[2]` patients diagnosed with schizophrenia or
schizoaffective disorder. There were `r ntottx[2]` 
(`r round(ntottx[2]/ntot[2], 2) * 100`%) patients randomly allocated to 
the treatment group, and `r ntotct[2]` 
(`r round(ntotct[2]/ntot[2], 2) * 100`%) to the placebo group. Patients 
in the treatment group received 1 out of 11 investigated antipsychotic 
drugs. Individual comparisons between drugs across studies indicated
marked differences between individual antipsychotics 
(`r parse_vrstat(vrstat[[2]])`; Figure \@ref(fig:fig6)).
The VR for ziprasidone, brexpiprazole, paliperidone, and olanzapine was 
smaller than 1. The VR for iloperidone was 1. The VR for risperidone, 
quetiapine, lurasidone, haloperidol, and sertindole was greater than 1.
Even though the variability for QTc prolongation was higher under 
treatment than under control, the difference did not reach statistical 
significance (`r parse_vrstat(vrstat[[2]])`; Figure \@ref(fig:fig5)).

```{r fig6capt}
fig6capt <- "Variability ratio for QTC prolongation for individual
antipsychotics. The forest plot shows the VR together with its 95% 
confidence interval (CI) for treatment versus placebo. All included studies
[@Daniel1999;@Correll2015;@Marder2007;
@Kane2015;@Potkin2007;@Nasrallah2013;@Potkin2003;@Arvanitis1997;
@Cutler2010;@Potkin2008;@Durgam2014;@Borison1996;@Borison1992;
@Kane2002;@Meltzer2011;@Patil2007;@Nakamura2009;@Harvey2013;
@Ogasa2013;@Potkin2015;@Zborowski1995;
@Lindenmayer2008;@Keck1998;@Garcia2009;@Beasley1996b;
@Zimbroff1997;@VanKammen1996]
are also listed in Table S1." 

```

```{r fig6, include = TRUE, fig.cap = fig6capt}
knitr::include_graphics(fig6)
```


# Discussion
## Summary
This study assessed the variability in the three major side effects of
antipsychotic treatment in schizophrenia spectrum disorders. We
focused on side effects because their occurrence has a great impact on
treatment adherence and physical health of patients, and clinical
experience suggests a potential to improve treatment allocation by
taking into account the variability in side effect occurrence. We also
know from clinical trials and meta-analyses that some antipsychotics
are more associated with specific side effects than others. For
example, clozapine and olanzapine are strongly associated with weight
gain,[@Huhn2019; @Homan2018; @Pillinger2020] QTc-time prolongation is
most distinct in sertindole and amisulpride,[@Huhn2019] and prolactin
level elevation in paliperidone and risperidone.[@Huhn2019] However,
these data cannot address the question whether there is variability in
subgroups or individual patients. Such side effect-by-subgroup or side
effect-by-patient interaction would be a prime example for the need of
a more stratified or personalized medicine, respectively, which
allocates treatments according to side effect profiles of subgroups or
individual patients. Overall, we found that the variability for weight
gain and prolactin elevation was indeed significantly increased in
patients who received treatment compared with those who received
placebo. For QTc prolongation, the increase was not
significant. Together, our results suggest that there is indeed marked
variability in the occurrence of side effects in antipsychotic
treatment. Variability also differed markedly between individual drugs.

## Reporting
Altogether we included `r sum(ntot)` patients from `r ntrialsincluded`
studies. The included studies provided data for treatment with 14 
antipsychotic drugs for weight gain, 13 antipsychotic drugs 
for hyperprolactinemia, and 11 antipsychotic drugs for 
QTc prolongation compared to placebo. Only for about 40% of studies 
included in a previous meta-analysis[@Huhn2019] variance data for at 
least one of the side effects of interest (weight gain, prolactin 
levels, QTc prolongation) were available. In about 62% of the studies 
included[@Huhn2019] incomplete data existed such that means were reported 
without a measure of variance. Although we did contact authors for missing 
data whenever possible, we received missing data only for three studies. 
In summary, consistent reporting of antipsychotic side effects, 
specifically with respect to variability measures, is currently missing 
in the literature and should be improved in future studies.

## Weight gain
Weight gain, especially for second generation
antipsychotics,[@Osborn2018] is a severe side effect that can
contribute to metabolic dysregulation. Importantly, every kg of weight
gain leads to a linear increase in the risk of cardiovascular
diseases,[@Willett1995] heart failure,[@Kenchaiah2002] and
diabetes.[@Cooper2016] Clozapine, olanzapine, zotepine, and sertindole
have the most severe impact in gaining weight. Some studies showed
that a lower BMI at baseline[@Gebhardt2009] and sex[@Najar2017] can
lead to more weight gain, whereas other studies found that male sex
and higher BMI at baseline are related to a higher risk of metabolic
disturbances.[@Pillinger2020] Our findings provide evidence that some
patients are indeed more susceptible to antipsychotic weight gain than
others, and that this susceptibility varies also between
medications. For example, for quetiapine we found that patients
differed in their susceptibility for weight gain while we did not find
such evidence for olanzapine and brexpiprazole, suggesting a potential
for stratified or personalized medicine for quetiapine but not
olanzapine and brexpiprazole. As antipsychotics in the treatment for
schizophrenia and related diseases is often recommended to be taken as
a relapse prevention for a longer period,[@Leucht2012a; @Homan2019h]
patients are likely to gain more weight during their treatment over
months and years. Together, this suggests that there is a potential to
improve long-term health and adherence by identifying the subgroups or
individual patients that are particularly prone to weight
gain. Preliminary evidence suggests that a dysregulated striatal
reward circuit contributes to this weight gain
susceptibility.[@Nielsen2016; @Homan2019k]

## Hyperprolactinemia
Prolactin level elevations occur in up to 70% of patients[@Inder2011]
under the treatment with antipsychotics. By blocking dopamine D2
receptors on lacotroph cells a disinhibition of the synthesis and
secretion of prolactin is observed.[@BenJonathan2001; @Bushe2010] This
can lead to both short- and long-term side effects with potentially
severe impact on patients' health. Typical short-term effects include
galactorrhea, gynecomastia, menstrual irregularities, and sexual
dysfunction; a typical long-term result is
osteoporosis,[@Mazziotti2018; @Byerly2007] and a potentially increased
risk in developing breast cancer in association with
hyperprolactinemia.[@Johnston2018; @George2020] Our findings suggest
that these risks may be particularly relevant for some patients but
not others, and more relevant for some antipsychotic drugs than
others. For example, a previous study found that prolactin level
elevations are more pronounced and more frequent in women than in
men.[@Veselinovic2011] In addition, some antipsychotics such as
amisulpride, risperidone, and paliperidone are linked to a greater
elevation of prolactin.[@Veselinovic2011; @Huhn2019] The striking
difference between risperidone (for which we did not find evidence for
significant variability in prolactin elevation) and paliperidone (for
which we did find such evidence) is puzzling as paliperidone is an
active metabolite of risperidone, and previous literature suggests
that paliperidone and risperidone lead to similar elevations in serum
prolactin concentrations [@Berwaerts2009]. A possible
explanation is that the level of prolactin can be highly variable
based on multiple biological and methodological factors such as
stress, diurnal variation and type of assay performed. However, future
studies should pay particular attention to differences in
susceptibility to prolactinemia between these two antipsychotics. In
summary, and in line with the weight gain findings, our findigns
suggest that there is a potential to improve long-term health and
antipsychotic adherence by identifying the subgroups or individual
patients that are particularly likely to develop prolactin elevations
under antipsychotic treatment.

## QTc prolongation
Prolongation of QTc is another important antipsychotic side effect as
cardiovascular diseases remain the most common cause of natural
mortality in schizophrenia spectrum disorders.[@Riordan2011] Users of
antipsychotic medication are reported to have higher rates of sudden
cardiac death than nonusers.[@Ray2009] Prolongation of QTc (longer
than 450 ms in men and longer than 470 ms in women, respectively, when
corrected with Bazetts Formula[@Bazett1997]) can contribute to
this.[@Funk2020] A prolongation of QTc can lead to *torsade de
pointes* and subsequently to sudden death.[@Glassman2001;
@Nielsen2011] The molecular pathway of this side effect is not
completely understood.[@Spellmann2018] It is known, however, that some
medications such as sertindole, amisulprid, and ziprasidone lead to
more QTc prolongation than others.[@Huhn2019] Here, we found increased
variability for some antipsychotics such as haloperidol in QTc
prolongation. However, altogether we did not find a statistically
significant increase in variability for QTc prolongation, potentially
because a smaller number of studies were available which decreased
statistical power.

## Limitations and strengths 
Our meta-analysis had some limitations. First, the occurrence of side
effects might be a dose-dependent effect which could reflect a
higher/different VR in some studies. Dose-dependent means and standard 
deviations are often missing but would be necessary to investigate 
dose-dependent effects. Second, for QTc, a reduced number of studies was
available, potentially reducing statistical power to detect a
significant variability increase. Third, our sample did not include 
pre-defined subgroups including patients with first-episode psychosis 
and treatment-resistant patients to create the most homogeneous sample 
possible. Finally, our method cannot determine
whether the increased variability is due to variability differences in
subgroups or individual patients.[@Cortes2019] The particular strength
of our study is that we included all available studies of antipsychotic
treatment in SSD reporting variability measures for side effects of
interest. To our knowledge, this is the first comprehensive study that
provides evidence for substantial variability in side effects.

## Conclusion
While we did not find convincing evidence that patients differed in
their susceptibility to QTc prolongation, we did find such evidence
for weight gain and prolactin elevation: for half of all
antipsychotics (7 out of 14) we can assume that subgroups of patients
or even individual patients would benefit from specific treatment
allocation through stratified or personalized medicine,
respectively. Such efforts in precision medicine might be crucial to
improve adherence and long-term health under antipsychotic treatment.

# Acknowledgements
The authors thank Majnu John, PhD, for advice on the analysis of the
current study and Ellen Ji, PhD, for her thoughtful comments on the
manuscript. These individuals received no additional compensation,
outside of their usual salary, for their contributions.

# Funding/Support
PH is supported by a NARSAD grant from the Brain & Behavior Research
Foundation (28445) and by a Research Grant from the Novartis Foundation
(20A058).

# Conflict of interest
In the last 3 years Dr. Leucht has received honoraria for service as a
consultant or adviser and/or for lectures from Angelini, Böhringer
Ingelheim, Geodon&Richter, Janssen, Johnson&Johnson, Lundbeck, LTS
Lohmann, MSD, Otsuka, Recordati, SanofiAventis, Sandoz, Sunovion,
TEVA. Dr. Kane reported grants from Otsuka, Lundbeck and Janssen, as
well as other from Alkermes, Allergan, Forum, Genentech, Lundbeck,
Intracellular Therapies, Janssen, Johnson & Johnson, Merck, Neurocrine,
Otsuka, Pierre Fabre, Reviva, Roche, Sunovion, Takeda, Teva, Vanguard
Research Group, and LB Pharmaceuticals outside of the submitted work. No
other disclosures were reported.

# Author contributions
MSN co-analyzed and interpreted the data, wrote the first draft of the
manuscript and revised the final manuscript. SH conceptualized the
study, wrote the primary analysis code and revised the final
manuscript. SV helped initiate the study and revised the final
manuscript. ES helped initiate the study and revised the final
manuscript. JMK helped initiate the study and revised the final
manuscript. MH collected the data and revised the final manuscript. SL
conceptualized the study and revised the final manuscript. PH initiated,
conceptualized and supervised the study, performed the statistical
analysis, and revised the final manuscript. All authors approved the
final version of the manuscript.

# References
<div id = "refs"></div> 

\onecolumn
\clearpage
\beginsupplement

# Supplementary Information
## Supplementary Table

```{r}
wglw <- wgl%>%
  dplyr::filter(index%in% c("weightsd", "prolactinsd", "qtcsd"))
wglw$bk <- ""
wglw$bk[!is.na(wglw$bibkey)] <- wglw$bibkey[!is.na(wglw$bibkey)]
wglw <- wglw%>%
  spread(key=index, value=val)
wglw$Weight <- ifelse(!is.na(wglw$weightsd), "Yes", "No")
wglw$Prolactin <- ifelse(!is.na(wglw$prolactinsd), "Yes", "No")
wglw$QTc <- ifelse(!is.na(wglw$qtcsd), "Yes", "No")
wglu <- wglw%>%
  #filter(drug=="Placebo")%>%
  dplyr::select(id, arm, bk, year, n, drug, Weight, Prolactin, QTc)%>%
  distinct(arm, .keep_all=TRUE)%>%
  mutate(Study=paste0(id, "[@", bk, "]"))%>%
  dplyr::select(Study, year, n, drug, Weight, Prolactin, QTc)%>%
  dplyr::arrange(-year) %>%
  dplyr::rename(
           `Study Arm`=Study,
           Year=year,
           `Sample Size`=n,
           Drug=drug
         )

#knitr::kable(wglu, format="markdown", longtable=TRUE,
#                  booktabs=TRUE,
#      caption = "All included study arms with references") 

#kableExtra::kable_styling(latex_options=c("hold_position",
#                                          "repeat_header"))

knitr::kable(wglu,
             format="pandoc",
             caption="All study arms with references")

```

\clearpage

## Supplementary Figures

```{r fig1capt}
fig1capt <- "Variability ratio for weight gain. The forest 
plot shows the VR together with its 95% confidence interval (CI) for 
treatment versus placebo. All included studies
[@Litman2016;@Garcia2009;@Clark1972;@Kahn2007;
@Ogasa2013;@Davidson2007;@Durgam2016;@Coppola2011;@Landbloom2017;
@Kinon2011;@VanKammen1996;@Arvanitis1997;@Ishigooka2018;
@Correll2015;@Zborowski1995;@Potkin2008;@Nakamura2009;
@Potkin2007;@Keck1998;@Cutler2010;@Clark1970;@Borison1996;@Ahmed2007;
@Durgam2015;@Meltzer2007;@Meltzer2004;@Kane2010;@Kane2002;@Hirayasu2010;
@Daniel1999;@Cooper2000;@Casey2008;@Bugarski2014;@Patil2007;
@Lieberman2015;@Potkin2003;@Durgam2014;@Nasrallah2013;@Marder2007;
@Borison1992;@Shen2014;@Kinoshita2016;@McEvoy2007;@Kane2007;@Harvey2013;
@Meltzer2011;@Canuso2010b;@Potkin2015;@Beasley1996a;@Canuso2010a;
@Beasley1996b;@Loebel2016;@Litman2014;@Kane2016;@Lindenmayer2008]
are also listed in Table S1."
```

```{r fig1, include = TRUE, fig.cap = fig1capt}
knitr::include_graphics(fig1)
```

```{r fig3capt}
fig3capt <- "Variability ratio for hyperprolactinemia. The forest 
plot shows the VR together with its 95% confidence interval (CI) for 
treatment versus placebo. All included studies
[@Kane2016;@Patil2007;@Kane2015;@Litman2016;
@Loebel2016;@Potkin2008;@Kahn2007;@Correll2015;@Kinoshita2016;
@VanKammen1996;@Small1997;@Kinon2011;@Potkin2003;@Arvanitis1997;
@Durgam2014;@Borison1996;@Schmidt2014;@Meltzer2007;@McEvoy2007;
@Hirayasu2010;@Casey2008;@Beasley1996a;@Marder2007;@Kane2007;
@Kane2010;@Durgam2016;@Lindenmayer2008;@Kane2002;@Ishigooka2018;
@Meltzer2011;@Durgam2015;@Cutler2010;@Harvey2013;@Nasrallah2013;
@Garcia2009;@Canuso2010a;@Nakamura2009;@Geffen2012;@Canuso2010b;
@Zimbroff1997;@Zborowski1995;@Beasley1996b;@Potkin2007;
@Potkin2015;@Davidson2007;@Ogasa2013;@Coppola2011;@Lieberman2015]
are also listed in Table S1." 
```

```{r fig3, include = TRUE, fig.cap = fig3capt}
knitr::include_graphics(fig3)
```

```{r fig5capt}
fig5capt <- "Variability ratio for QTc prolongation. The forest 
plot shows the VR together with its 95% confidence interval (CI) for 
treatment versus placebo. All included studies 
[@Daniel1999;@Correll2015;@Marder2007;
@Kane2015;@Potkin2007;@Nasrallah2013;@Potkin2003;@Arvanitis1997;
@Cutler2010;@Potkin2008;@Durgam2014;@Borison1996;@Borison1992;
@Kane2002;@Meltzer2011;@Patil2007;@Nakamura2009;@Harvey2013;
@Ogasa2013;@Potkin2015;@Zborowski1995;
@Lindenmayer2008;@Keck1998;@Garcia2009;@Beasley1996b;
@Zimbroff1997;@VanKammen1996]
are also listed in Table S1." 

```

```{r fig5, include = TRUE, fig.cap = fig5capt}
knitr::include_graphics(fig5)
```
